---
title: "Analysis of vowel categories across corpus, gender, and consonantal context"
author: "Annie Helms (annie_helms@berkeley.edu)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "cairo_pdf")
options(max.print = 3000)
```

Library packages

```{r message = FALSE, warning = FALSE}
library(r2glmm)
library(dplyr)
library(ggplot2)
library(lmerTest) #mixed effects linear regression
library(ggsignif) # helps plot significant relations in ggplot2
library(kableExtra) # formats dataframes in html output
library(BSDA) # manual t-test with means and sd info
library(emmeans) # tukey post-hoc tests
```

### Upload data set

```{r message = FALSE, echo = FALSE}
setwd("/Users/Annie/Box Sync/PhD/github/Bay_Area_Spanish/")
```

Upload productions of /a/ across stress, corpus, consonantal context, and gender.

```{r}
cons_model = read.csv("plot_model/cons_cont.csv", 
                  stringsAsFactors = TRUE)
head(cons_model, 5) %>%
  kable() %>% # piped to allow for kable formatting
  kable_styling(bootstrap_options = c("striped", # alternates row color in table
                                      "hover", # rows are highlighted when hovered over
                                      "condensed", # row height is condensed
                                      "responsive")) %>% # horizontal scrolling across table
  scroll_box(width = "100%")
```

Models first. Three models: duration (normalized), F1 and F2.

```{r}
# submit to mixed effects linear regression model
dur_model = lmer(Dur_norm ~ Corpus * Vowel * prev_ph + (1|Participant),
                   data = cons_model, REML = FALSE)
summary(dur_model)
```

Post-hoc tests:

```{r message = FALSE}
emmeans(dur_model, list(pairwise ~ prev_ph * Corpus), adjust="tukey")
```

According to the output above, there is no significant effect of consonantal context on duration across the corpora.

```{r}
# submit to mixed effects linear regression model
f1_model = lmer(F1.50_norm ~ Corpus * prev_ph * Vowel + (1|Participant),
                   data = cons_model, REML = FALSE)
summary(f1_model)
```

According to the output above, there is no significant difference in F1 due to consonantal context across the corpora.

```{r}
# submit to mixed effects linear regression model
f2_model = lmer(F2.50_norm ~ Corpus * prev_ph * Vowel + (1|Participant),
                   data = cons_model, REML = FALSE)
summary(f2_model)
```

According to the output of the model above, the interaction of previous phone and corpus is not significantly, meaning that consonantal environment has the same effect on vowels in both corpora.



