---
title: "Analysis of vowel space compression across gender and corpus"
author: "Annie Helms (annie_helms@berkeley.edu)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "cairo_pdf")
options(max.print = 1500)
```

Library packages

```{r message = FALSE, warning = FALSE}
library(r2glmm)
library(dplyr)
library(ggplot2)
library(lmerTest) #mixed effects linear regression
library(ggsignif) # helps plot significant relations in ggplot2
library(kableExtra) # formats dataframes in html output
library(BSDA) # manual t-test with means and sd info
library(emmeans) # tukey post-hoc tests
```

### Upload data set

```{r message = FALSE, echo = FALSE}
setwd("/Users/Annie/Box Sync/PhD/github/Bay_Area_Spanish/")
```

Upload productions of vowels across corpus and gender.

```{r}
area_model = read.csv("plot_model/area_model.csv", 
                  stringsAsFactors = TRUE)
head(area_model, 5) %>%
  kable() %>% # piped to allow for kable formatting
  kable_styling(bootstrap_options = c("striped", # alternates row color in table
                                      "hover", # rows are highlighted when hovered over
                                      "condensed", # row height is condensed
                                      "responsive")) %>% # horizontal scrolling across table
  scroll_box(width = "100%")
```

Create smaller data set and calculate area by speaker.

```{r}
area_speaker = area_model %>%
  group_by(Participant, Gender, Corpus) %>% 
  summarize(Area = mean(Area))
```

Run regression model:

```{r}
# submit to fixed effects linear regression model
area_speaker_model = glm(Area ~ Corpus * Gender,
                   data = area_speaker)
summary(area_speaker_model)
```

According to the model output above, there is no significant difference between vowel space area across gender and corpus.

Now look at dispersion.

```{r}
disp_model = read.csv("plot_model/disp_model.csv", 
                  stringsAsFactors = TRUE)
head(disp_model, 5) %>%
  kable() %>% # piped to allow for kable formatting
  kable_styling(bootstrap_options = c("striped", # alternates row color in table
                                      "hover", # rows are highlighted when hovered over
                                      "condensed", # row height is condensed
                                      "responsive")) %>% # horizontal scrolling across table
  scroll_box(width = "100%")
```

Create smaller data set and calculate dispersion by speaker.

```{r}
disp_speaker = disp_model %>%
  group_by(Participant, Gender, Corpus) %>% 
  summarize(Disp = mean(Dispersion))
```

Now submit to fixed effects linear regression model.

```{r}
# submit to fixed effects linear regression model
disp_speaker_model = glm(Disp ~ Corpus * Gender,
                   data = disp_speaker)
summary(disp_speaker_model)
```

According to the model output above, there is also no significant difference in vowel dispersion across gender and corpus.
