---
title: "Vowel contours for Spanish data"
author: "Annie Helms (annie_helms@berkeley.edu)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
setwd("/Users/Annie/Box Sync/PhD/github/Bay_Area_Spanish/")
```

Following the methodology laid out in Fruewahld (2010) (https://www.ling.upenn.edu/~joseff/papers/fruehwald_ssanova.pdf), install the package `gss` to fit the Smoothing Spline ANOVA (SS-ANOVA) to the resulting formant data. 

```{r warning = FALSE, message = FALSE}
# run this cell to install package `gss`
# install.packages("gss")
```
Install necessary libraries:

```{r message = FALSE, warning = FALSE}
library(dplyr) # manipulating data frames
library(ggplot2) # making visualizations
library(kableExtra) # formats dataframes in html output
library(gss) #SS-ANOVA
```

Upload the .csv file with the normalized formant measurements, extracted every 2ms. The data needs to be in a format where `F1` is a column, as well as `RTime` (scaled time variable measuring 2ms intervals).

```{r}
formants = read.csv("data/contour_norm.csv", 
                  stringsAsFactors = TRUE)
head(formants, 5) %>%
  kable() %>% # piped to allow for kable formatting
  kable_styling(bootstrap_options = c("striped", # alternates row color in table
                                      "hover", # rows are highlighted when hovered over
                                      "condensed", # row height is condensed
                                      "responsive")) %>% # horizontal scrolling across table
  scroll_box(width = "100%")
```

Since this data set contains Spanish vowels, create subsets of the data for each vowel in question. From these subsets, randomly sample 100 productions.

```{r}
a = formants %>% 
  filter(Vowel=="a") %>% 
  filter(unique %in% sample(levels(unique),100))

e = formants %>% 
  filter(Vowel=="e")%>% 
  filter(unique %in% sample(levels(unique),100))

i = formants %>% 
  filter(Vowel=="i")%>% 
  filter(unique %in% sample(levels(unique),100))

o = formants %>% 
  filter(Vowel=="o") %>% 
  filter(unique %in% sample(levels(unique),100))

u = formants %>% 
  filter(Vowel=="u") %>% 
  filter(unique %in% sample(levels(unique),100))
```

Make a preliminary plots of /a/ productions, and visually see where formant tracking errors may be. the factor `unique` is used to group together trajectories of the same token by the same speaker.

```{r}
a_track = ggplot(data = a,
                 aes(x = RTime,
                     group = unique,
                     color = stress)) +
  geom_line(aes(y = F1_norm),
            alpha = 0.8) +
  geom_line(aes(y = F2_norm),
            alpha = 0.8) +
  theme_classic()
a_track
```

Now to fit the data with SS-ANOVA.

```{r}
f1_a = ssanova(F1_norm ~ stress * RTime, data = a)
f2_a = ssanova(F2_norm ~ stress * RTime, data = a)
```

To test the fit of the data, generate a dummy set of 100 evenly spaced points across our factor `RTime` for each level of `stress`.

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed","unstressed"))
```

Store the results of the predict function (comparing model to dummy set).

```{r}
grid$F1.Fit = predict(f1_a, newdata = grid, se = T)$fit
grid$F1.SE = predict(f1_a, newdata = grid, se = T)$se.fit

grid$F2.Fit = predict(f2_a, newdata = grid, se = T)$fit
grid$F2.SE = predict(f2_a, newdata = grid, se = T)$se.fit
```

Plot the data stored in grid, with 95% confidence interval:

```{r}
ss_a = ggplot(data = grid,
              aes(x = RTime,
                  color = stress,
                  group = stress)) +
  geom_line(aes(y = F1.Fit),
            alpha = 1,
            color = "grey20")+ 
  geom_line(aes(y = F2.Fit),
            alpha = 1,
            color = "grey20") +
  geom_ribbon(aes(ymin = F1.Fit-(1.96*F1.SE),
                  ymax = F1.Fit+(1.96*F1.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") + 
  geom_ribbon(aes(ymin = F2.Fit-(1.96*F2.SE),
                  ymax = F2.Fit+(1.96*F2.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") +
  labs(y = "Normalized Frequency",
       x = "Scaled Time",
       title = "Formant Trajectories of Spanish /a/") +
  guides(fill = guide_legend(title = "Stress")) +
  scale_fill_hue(labels = c("Stressed", "Unstressed")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "right")
ss_a
```

```{r echo = FALSE, eval = FALSE}
# save image to pdf
cairo_pdf("figs/contour_a.pdf", family = "Charis SIL", width = 7, height = 4)
ss_a
dev.off()
```

Now to measure statistically whether stress significantly impacts the formant shape.

```{r}
## create prediction grid
grid = expand.grid(RTime = seq(0,1,length = 100),stress = c("stressed","unstressed"))

## Predict for certain terms
## The "inc" argument to predict is key here
grid$F1.Fit = predict(f1_a, grid, se = T,inc = c("1","RTime"))$fit
grid$F1.SE = predict(f1_a, grid, se = T,inc = c("1","RTime"))$se.fit
grid$F2.Fit = predict(f2_a, grid, se = T,inc = c("1","RTime"))$fit
grid$F2.SE = predict(f2_a, grid, se = T,inc = c("1","RTime"))$se.fit
```

Now we can plot this underlying form, without effects of stress.

```{r}
bare_a = ggplot(grid, aes(x = RTime)) +
  geom_line(aes(y= F1.Fit)) +
  geom_line(aes(y= F2.Fit)) +
  geom_ribbon(aes(ymax= F1.Fit +(1.96*F1.SE),
                  ymin= F1.Fit -(1.96*F1.SE)),
              alpha = 0.5) +
  geom_ribbon(aes(ymax= F2.Fit +(1.96*F2.SE),
                  ymin= F2.Fit -(1.96*F2.SE)),
              alpha = 0.5) +
  labs(y = "Normalized Frequency",
       x = "Scaled Time",
       title = "Underlying Formant Trajectory of Spanish /a/") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "right") +
  ylim(c(.4,1.6))
bare_a
```

Use the following code to predict group and interaction effects for F1:

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed", "unstressed"))
grid$Fit = predict(f1_a, grid, se = T, inc = c("stress", "stress:RTime"))$fit
grid$SE = predict(f1_a, grid, se = T, inc = c("stress", "stress:RTime"))$se.fit
```

```{r}
eff = ggplot(grid, aes(x = RTime)) +
  geom_line(aes(y = Fit))+
  geom_ribbon(aes(ymax = Fit+(1.96*SE),
                  ymin = Fit-(1.96*SE)),
              alpha = 0.5) +
  facet_wrap(~stress)+
  geom_abline(slope = 0,intercept = 0,lty = 2)+
  ylab("Difference in Hz")
eff
```

For F2:

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed", "unstressed"))
grid$Fit = predict(f2_a, grid, se = T, inc = c("stress", "stress:RTime"))$fit
grid$SE = predict(f2_a, grid, se = T, inc = c("stress", "stress:RTime"))$se.fit
```

Now to plot:

```{r}
eff = ggplot(grid, aes(x = RTime)) +
  geom_line(aes(y = Fit))+
  geom_ribbon(aes(ymax = Fit+(1.96*SE),
                  ymin = Fit-(1.96*SE)),
              alpha = 0.5) +
  facet_wrap(~stress)+
  geom_abline(slope = 0,intercept = 0,lty = 2)+
  ylab("Difference in Hz")
eff
```

Confidence intervals (95%) are used, and since 0 difference is included, the group effect is not significant for F1 or F2.

Predicting group and interaction effects:

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed", "unstressed"))
grid$Fit <- predict(f1_a, grid, se = T, inc = c("stress", "stress:RTime"))$fit
grid$SE <- predict(f1_a, grid, se = T, inc = c("stress", "stress:RTime"))$se.fit
```

Plot:

```{r}
eff = ggplot(grid, aes(x = RTime)) +
  geom_line(aes(y = Fit))+ 
  geom_ribbon(aes(ymax = Fit+(1.96*SE),
                  ymin = Fit-(1.96*SE)),
              alpha = 0.5) +
  facet_wrap(~stress)+ 
  geom_abline(slope = 0, intercept = 0,lty = 2)+
  ylab("Difference in Normalized Frequency")
eff
```

From above, F1 is significantly different in different stress environments until about 60% into the vowel production.

Now for F2.

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed", "unstressed"))
grid$Fit = predict(f2_a, grid, se = T, inc = c("stress", "stress:RTime"))$fit
grid$SE = predict(f2_a, grid, se = T, inc = c("stress", "stress:RTime"))$se.fit
```

Plot:

```{r}
eff = ggplot(grid, aes(x = RTime)) +
  geom_line(aes(y = Fit))+ 
  geom_ribbon(aes(ymax = Fit+(1.96*SE),
                  ymin = Fit-(1.96*SE)),
              alpha = 0.5) +
  facet_wrap(~stress)+ 
  geom_abline(slope = 0, intercept = 0,lty = 2)+
  ylab("Difference in Normalized Frequency")
eff
```

From above, stress does not significantly effect F2 production until above 60% into the vowel production.

Fit and plot for other vowels.

```{r}
f1_e = ssanova(F1_norm ~ stress * RTime, data = e)
f2_e = ssanova(F2_norm ~ stress * RTime, data = e)
```

To test the fit of the data, generate a dummy set of 100 evenly spaced points across our factor `RTime` for each level of `stress`.

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed","unstressed"))
```

Store the results of the predict function (comparing model to dummy set).

```{r}
grid$F1.Fit = predict(f1_e, newdata = grid, se = T)$fit
grid$F1.SE = predict(f1_e, newdata = grid, se = T)$se.fit

grid$F2.Fit = predict(f2_e, newdata = grid, se = T)$fit
grid$F2.SE = predict(f2_e, newdata = grid, se = T)$se.fit
```

Plot the data stored in grid, with 95% confidence interval:

```{r}
ss_e = ggplot(data = grid,
              aes(x = RTime,
                  color = stress,
                  group = stress)) +
  geom_line(aes(y = F1.Fit),
            alpha = 1,
            color = "grey20")+ 
  geom_line(aes(y = F2.Fit),
            alpha = 1,
            color = "grey20") +
  geom_ribbon(aes(ymin = F1.Fit-(1.96*F1.SE),
                  ymax = F1.Fit+(1.96*F1.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") + 
  geom_ribbon(aes(ymin = F2.Fit-(1.96*F2.SE),
                  ymax = F2.Fit+(1.96*F2.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") +
  labs(y = "Normalized Frequency",
       x = "Scaled Time",
       title = "Formant Trajectories of Spanish /e/") +
  guides(fill = guide_legend(title = "Stress")) +
  scale_fill_hue(labels = c("Stressed", "Unstressed")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "right")
ss_e
```

```{r echo = FALSE, eval = FALSE}
# save image to pdf
cairo_pdf("figs/contour_e.pdf", family = "Charis SIL", width = 7, height = 4)
ss_e
dev.off()
```

Now /o/:

```{r}
f1_o = ssanova(F1_norm ~ stress * RTime, data = o)
f2_o = ssanova(F2_norm ~ stress * RTime, data = o)
```

To test the fit of the data, generate a dummy set of 100 evenly spaced points across our factor `RTime` for each level of `stress`.

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed","unstressed"))
```

Store the results of the predict function (comparing model to dummy set).

```{r}
grid$F1.Fit = predict(f1_o, newdata = grid, se = T)$fit
grid$F1.SE = predict(f1_o, newdata = grid, se = T)$se.fit

grid$F2.Fit = predict(f2_o, newdata = grid, se = T)$fit
grid$F2.SE = predict(f2_o, newdata = grid, se = T)$se.fit
```

Plot the data stored in grid, with 95% confidence interval:

```{r}
ss_o = ggplot(data = grid,
              aes(x = RTime,
                  color = stress,
                  group = stress)) +
  geom_line(aes(y = F1.Fit),
            alpha = 1,
            color = "grey20")+ 
  geom_line(aes(y = F2.Fit),
            alpha = 1,
            color = "grey20") +
  geom_ribbon(aes(ymin = F1.Fit-(1.96*F1.SE),
                  ymax = F1.Fit+(1.96*F1.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") + 
  geom_ribbon(aes(ymin = F2.Fit-(1.96*F2.SE),
                  ymax = F2.Fit+(1.96*F2.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") +
  labs(y = "Normalized Frequency",
       x = "Scaled Time",
       title = "Formant Trajectories of Spanish /o/") +
  guides(fill = guide_legend(title = "Stress")) +
  scale_fill_hue(labels = c("Stressed", "Unstressed")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "right")
ss_o
```


```{r echo = FALSE, eval = FALSE}
# save image to pdf
cairo_pdf("figs/contour_o.pdf", family = "Charis SIL", width = 7, height = 4)
ss_o
dev.off()
```

Now /u/:

```{r}
f1_u = ssanova(F1_norm ~ stress * RTime, data = u)
f2_u = ssanova(F2_norm ~ stress * RTime, data = u)
```

To test the fit of the data, generate a dummy set of 100 evenly spaced points across our factor `RTime` for each level of `stress`.

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed","unstressed"))
```

Store the results of the predict function (comparing model to dummy set).

```{r}
grid$F1.Fit = predict(f1_u, newdata = grid, se = T)$fit
grid$F1.SE = predict(f1_u, newdata = grid, se = T)$se.fit

grid$F2.Fit = predict(f2_u, newdata = grid, se = T)$fit
grid$F2.SE = predict(f2_u, newdata = grid, se = T)$se.fit
```

Plot the data stored in grid, with 95% confidence interval:

```{r}
ss_u = ggplot(data = grid,
              aes(x = RTime,
                  color = stress,
                  group = stress)) +
  geom_line(aes(y = F1.Fit),
            alpha = 1,
            color = "grey20")+ 
  geom_line(aes(y = F2.Fit),
            alpha = 1,
            color = "grey20") +
  geom_ribbon(aes(ymin = F1.Fit-(1.96*F1.SE),
                  ymax = F1.Fit+(1.96*F1.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") + 
  geom_ribbon(aes(ymin = F2.Fit-(1.96*F2.SE),
                  ymax = F2.Fit+(1.96*F2.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") +
  labs(y = "Normalized Frequency",
       x = "Scaled Time",
       title = "Formant Trajectories of Spanish /u/") +
  guides(fill = guide_legend(title = "Stress")) +
  scale_fill_hue(labels = c("Stressed", "Unstressed")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "right")
ss_u
```


```{r echo = FALSE, eval = FALSE}
# save image to pdf
cairo_pdf("figs/contour_u.pdf", family = "Charis SIL", width = 7, height = 4)
ss_u
dev.off()
```

Now /i/:

```{r}
f1_i = ssanova(F1_norm ~ stress * RTime, data = i)
f2_i = ssanova(F2_norm ~ stress * RTime, data = i)
```

To test the fit of the data, generate a dummy set of 100 evenly spaced points across our factor `RTime` for each level of `stress`.

```{r}
grid = expand.grid(RTime = seq(0, 1, length = 100), stress = c("stressed","unstressed"))
```

Store the results of the predict function (comparing model to dummy set).

```{r}
grid$F1.Fit = predict(f1_i, newdata = grid, se = T)$fit
grid$F1.SE = predict(f1_i, newdata = grid, se = T)$se.fit

grid$F2.Fit = predict(f2_i, newdata = grid, se = T)$fit
grid$F2.SE = predict(f2_i, newdata = grid, se = T)$se.fit
```

Plot the data stored in grid, with 95% confidence interval:

```{r}
ss_i = ggplot(data = grid,
              aes(x = RTime,
                  color = stress,
                  group = stress)) +
  geom_line(aes(y = F1.Fit),
            alpha = 1,
            color = "grey20")+ 
  geom_line(aes(y = F2.Fit),
            alpha = 1,
            color = "grey20") +
  geom_ribbon(aes(ymin = F1.Fit-(1.96*F1.SE),
                  ymax = F1.Fit+(1.96*F1.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") + 
  geom_ribbon(aes(ymin = F2.Fit-(1.96*F2.SE),
                  ymax = F2.Fit+(1.96*F2.SE),
                  fill = stress),
              alpha = 0.5,
              color = "NA") +
  labs(y = "Normalized Frequency",
       x = "Scaled Time",
       title = "Formant Trajectories of Spanish /i/") +
  guides(fill = guide_legend(title = "Stress")) +
  scale_fill_hue(labels = c("Stressed", "Unstressed")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "right")
ss_i
```


```{r echo = FALSE, eval = FALSE}
# save image to pdf
cairo_pdf("figs/contour_i.pdf", family = "Charis SIL", width = 7, height = 4)
ss_i
dev.off()
```


